---
title: "Exploratory Analysis using credit dataset"
author: "Himansu Sahoo"
date: "February 2, 2016"
output: html_document
---
This is the example showing how to do exploratory analysis using German Credit dataset.
In the next examples, we will apply decision tree algorithms.

### Step 1 : Getting Data into R

```{r}
raw_data <- read.csv(file="../datasets/credit.csv", na.strings=c(""," ", "NA", "NAN"))
```

#### Checking the dataset 

```{r}
#class(raw_data) # data.frame
dim(raw_data) # 1000 21
#nrow(raw_data); ncol(raw_data)
#names(raw_data)
colnames(raw_data) # same as above
str(raw_data) # will show class of each variable
#summary(raw_data) # five number summary of the dataset
```

The dataset contains 1000 observations and 21 variables.
rows are the observations, columns are the variables.
The number 21 variable (default) is the response variable.

#### Print some rows
```{r}
#head(raw_data) # will print first 6 rows
#tail(raw_data) # will print last 6 rows
#head(raw_data, n=50) # will print first 50 rows
```

#### Checking the response variable

**default** is the response variable. Check different properties of the response variable.

```{r}
# get the column number of the default variable
print(which(colnames(raw_data) %in% "default"))
#print(which(colnames(raw_data) %in% c("default"))) # same as above
#print(which(colnames(raw_data) == "default")) # same as above

#class(raw_data$default) # integer
str(raw_data$default) # int [1:1000] 1 2 1 1 2 1 1 1 1 2 ...
#summary(raw_data$default)
#unique(raw_data$default) # number of unique value 1 2
```

The response variable **default** is of class integer with only two unique values 1, 2. covert this to a factor variable with two levels.


```{r}
#raw_data$default <- as.factor(raw_data$default)
# give them valid R names
raw_data$default <- as.factor(ifelse(raw_data$default == 1, "NO", "YES"))

#class(raw_data$default) # factor
str(raw_data$default) ##  Factor w/ 2 levels "1","2": 1 2 1 1 2 1 1 1 1 2 ...
#levels(raw_data$default) # "1" "2"
table(raw_data$default)
#summary(raw_data$default) # same as above
prop.table(table(raw_data$default))
```

#### Check NA's in the dataset

Looking at `summary(raw_data)` doesn't show any NA values.
```{r}
#summary(raw_data)
colSums(is.na(raw_data))
```










